Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2021-02-02 manuscript/draft_rsfmri.Rmd

---
title: "Effects of bariatric surgery and weight loss on resting state functional connectivity"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Kristin Prehn
- Jürgen Ordemann
- Agnes Floel
- A. Veronica Witte [1][2][3]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
    extra_dependencies: "subfig"
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

```{r "To do", include=FALSE}

# [ ] Evaluate SPM model comparison (more activation in SPM..)
# [ ] Discussion
# [ ] brain images mit nilearn plotting tools einbinden
# [ ] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [ ] repeat analysis with exclusion of 10% of worst data ??
# [ ] Network tests
# [ ] Swap mFD with logmFD ??

@@

#library(lsr)
library(wesanderson) # colour palette
#library(haven)

@@

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = '') # hide NA in tables
knitr::knit_engines$get()
# set working directory of knit (RMarkdown) to directory of Rproj
#opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
#setwd(knitr::opts_knit$get("root.dir"))

@@

\newpage
# Introduction
Obesity is a worldwide health issue, enchaining huge personal and societal costs. Excess amount of fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life.
Conservative treatment options including dietary modification, physical activity and behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI (> 35 kg/m2). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status.
Roux-en-Y gastric bypass (RYGB) is the most effective surgical procedure [@Berthoud_2011]. Here, a small pouch is formed from the proximal stomach and connected to the jejunum, disconnecting the larger part of the stomach and the proximal part of the small intestine intact from the digestive flow. Other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) reduce stomach volume, but leave the small intestine unchanged [@Mulla_2017].
RYGB-induced weight loss is in part mediated by malabsorption of certain nutrients, especially fat, and reduced digestion efficiency [@Berthoud_2011] while the role of increased energy expenditure after RYGB is still unclear [@Yarmush_2017;@Lamarca_2019]. Most importantly, RYGB alters the signaling cascades between the gut, fat tissue and the brain and thereby induces profound changes in the regulation of food intake and energy homeostasis [@Berthoud_2011]. RYBG-patients frequently report altered smell and taste perception, i.e. increased sensitivity to sweet taste, show less desire to consume high-calorie and high-fat foods and actually reduce intake of such food items [@Mulla_2017, @Berthoud_2011].  These changes in appetitive and reward signaling likely relate to altered gut-brain communication, mediated by afferent vagal nerve signaling and gut hormones like GLP-1, PYY and ghrelin [ @Brutman_2019].
Yet, precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated.
Resting state functional magnetic resonance imaging (rsfMRI) is a powerful technique to investigate the dynamic organization of the brain via the correlation of spontaneous neural activity. Anatomically distinct brain regions show functional connectivity (FC), i.e. correlated neural activity over time, in the absence of a cognitive task. Crucially, these brain networks correspond to networks activated during specific cognitive processes, e.g. visual, auditory and somatosensory processing, or higher-order cognitive functions, i.e. attention and executive control [@Smith_2009], and can thus be seen as default configuration of the brain. In the context of bariatric surgery, the default mode network, a higher-order network, involved in interoception and governing shifts between external-internal processes, and the reward network, processing hedonic value and internal motivation, are promising candidates to mediate altered gut-brain signaling.
The default mode network has been discovered as a “task-negative” network, and is involved in a variety of largely internally-directed cognitive functions including interoception, mind wandering and memory. Its core hubs are the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015].
Higher BMI and obesity have been associated with differences in DMN FC in several resting state and task-based fMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015, @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018].
Despite the large heterogenity and low median power, these findings might be summarized as a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks. This has been interpreted as impaired interoceptive processing failing to adequately balance internal and external behavioral cues.
Longitudinal studies largely support this view. Increased activation of the DMN in a food-viewing paradigm, i.e. insufficient down-regulation of DMN, has been shown to correlate with appetite measures after an overnight fast [@Tregellas_2011, @Wijngaarden_2015], an effect that could be partially reversed by an exercise intervention [@McFadden_2013]. In a bariatric surgery study, [@Li_2018] reported a reduction of local FC in mid-line DMN regions in surgical patients compared to an obese control group. Simultaneously, PCC and medial prefrontal cortex were more strongly connected to dorsolateral prefrontal cortex, a region implicated in cognitive control. Another study in bariatric surgery patients showed that heightened connectivity of the PCC with anterior superior temporal gyrus (STG) before the surgery decreased over the course of one year [@Olivo_2017]. Similarly, the profile of DMN connectivity with frontal regions, including ACC, frontal superior gyrus and the right OFC,  was similar in patients after bariatric surgery and normal weight controls, but dissimilar to that of obese controls [@Frank_2013].
Another important brain network in the context of obesity is the reward network, comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of reward network regions during anticipation of (high-caloric) food cues, and in contrast, reduced reward region activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020;]. Food cue hyperactivation has been identified as a main neural vulnerability factor predicting future weight gain [@Stice_2015]. Resting state fMRI studies have partly mirrored these results, showing increased local FC of reward network regions, i.e. Nacc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. Imaging studies in bariatric surgery patients provided first evidence that hyperconnectivity between reward network regions (e.g. putamen, OFC) might be normalized to some extent by the surgery [@Duan_2020; @Wiemerslage_2016]. Possible, the reconfiguration of the fronto-striatal brain networks might be related to altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013, but @Zoon_2018] .
Thus, while there is some evidence hinting to a role for DMN and reward network connectivity in altered gut-brain communication after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohort of patients, often without adequate control group, and employed exploratory statistics [@George_2016].
In this study, we aimed to extend the current literature and investigate differences of DMN and reward network connectivity before and after bariatric surgery in preregistered analyses and in comparison with a contemporaneously measured waiting-control group. Moreover, we explored whether the changes in functional connectivity were associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential dose-response relationship.
Participants were measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we took into account possible non-linear time courses in increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017].

@@

```{r "tableSample", echo=FALSE}
#tableSample <- mk_SampleTable(final)
tableSample <- readRDS(file = "../report/tab/tableSample.rds")

tab <-
  knitr::kable(
    tableSample,
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "Distribution of data points at months after intervention")
tab
```

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité Berlin. For more details see @Prehn_2020. We used all data acquired until April 2019.  
The study protocol was in accordance with the Helsinki Declaration and approved by the local ethics committee. The study was registered at clinicaltrials.gov as NCT01554228. We preregistered the present resting state analyses on the [OSF](https://osf.io/yp42s). We made [additional changes](https://osf.io/59bh7/) to the preregistration after preprocessing as we realized some aspects of the analysis were inadequately described in the first preregistration.   
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->
Participants were recruited from the Center for Bariatric and Metabolic Surgery. Inclusion criteria were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants were aged between $18$ and $70$ and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$).
In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see section \@ref{sec:Anatomical Quality control}).
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)). In more detail, `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point. In the intervention group, data is not missing completely at random, as the pre-surgery timepoint MRI could not be obtained in some patients due to the limited scanner bore.
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "../report/dfBL.rds")

@@

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
<!-- BASELINE CHARACTERISTICS -->
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).

```{r tableDescr, include=FALSE}
# table for descriptive statistics
tableDescr <- readRDS(file = "../report/tab/tableDescr.rds")

knitr::kable(
    tableDescr,
    row.names = FALSE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "BMI and mean FD of participants acquired in each condition and time point")
```

@@

```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
readRDS(file = "../report/fig/fig_BMIdescr.rds")
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropocentric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

@@

#### Functional Quality control
Resting state fMRI data quality control was performed according to the protocol by @Ciric_2018.
As the expected average head motion was high in this sample, it was difficult to determine an appropriate exclusion threshold which would not lead to the exclusion of a high number of individuals [@Beyer_2020]. Thus, we did not exclude any participants based on mean framewise displacement (FD) estimates beyond those who failed the anatomical preprocessing. We applied best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. In addition, for models showing a significant time-by-group interaction, we performed a sensitivity analysis excluding the 10% data sets with highest mean FD (mFD).
```{r carpetplot, out.width="49%", out.height="30%", fig.cap="Carpet plot for quality control", fig.subcap=c("Example 1 with strong amount of structured noise", "Example 2 with less amount of structured noise"), fig.show='hold',fig.align='center', echo=FALSE}
carpets=c("../report/carpetplotadi006fu2.jpeg","../report/carpetplotadi16fu.jpeg")
knitr::include_graphics(carpets)
```
On the single-subject level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref(fig:carpetplot) shows an exemplary carpet plot with (from top to bottom): percent outliers defined by AFNI, mFD, `OutlierCount`, DVARS (spatial root mean square of the data after temporal differencing),voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data (red: GM voxels, green: WM voxels, blue: ventricle, green: cerebellum). All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.
We evaluated mFD-FC relationships by computing a functional connectome based on [@Power_2012]. In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we correlated mean FD and QC values for each node across participants, and plotted the resulting mFD-FC correlation values in histogram (see Figure \@ref{exampleFC}. We reported the mFD-FC correlation per preprocessing scheme.
Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward network over all participants and timepoints, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with N=999. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of p_FWE < 0.05. We extracted the average connectivity from these masks for GSR regressed data.

<!-- *In the preregistration we wrote: "We will use the average of connectivity values within masks of the respective networks as aggregate network connectivity measure. The mask will include all voxels surviving FWE-correction of p<0.05 in a one-sample t-test of the individual connectivity maps across all participants and time points, calculated with the modified SwE option in the SwE toolbox"* -->

@@

```{r "qa check DMN model age,sex", echo=FALSE}
diagnostics.plot(R0)
ranef.diagn.plot(R0)

>>>>

---
title: "Effects of bariatric surgery and weight loss on resting state functional connectivity"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Kristin Prehn
- Jürgen Ordemann
- Agnes Floel
- A. Veronica Witte [1][2][3]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

```{r "To do", include=FALSE}
# [ ] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [ ] Swap mFD with logmFD ??
```

@@

#library(lsr)
library(wesanderson) # color palette
#library(haven)

@@

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = ' ') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
#opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
#setwd(knitr::opts_knit$get("root.dir"))

@@

Obesity imposes serious health risks and is consistently associated with differences in brain activity, including resting-state functional connectivity which can be effectively treated by bariatric surgery. 
In this study, we investigated the effects of bariatric surgery and weight loss indicated by BMI on functional connectivity in the reward network and default mode network (DMN). Therefore, 33 bariatric surgery patients and 15 obese waiting-list control patients underwent MRI at baseline (before surgery/waiting period), after 6 and 12 months.
To adequately control for motion artifacts shown to be an issue in obesity studies, we applied different denoising pipelines and tried to disentangle the effects of weight loss and head motion in the second level analyses.
Although we did not find the proposed group-by-time interaction effect, head motion and BMI were significantly associated with functional connectivity. When adjusting for differences in head motion within and between individuals using mean framewise displacement, both, BMI decrease and higher average head motion, resulted in stronger functional connectivity in medial posterior frontal regions within the reward network.
While these findings could imply that severe weight loss changes reward-related neural processing in obese patients, the overlapping effects of BMI change and head motion on FC in this brain area, together with previous findings showing a strong covariance between BMI and head motion renders definite conclusions difficult.
Future studies with a more rigorous control of within-scanner head motion during data acquisition are warranted to further understand the effects of obesity surgery on functional connectivity measures.

\newpage
# Introduction
Obesity is a worldwide health issue, enchaining huge personal and societal costs. Excess amount of fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life.
Conservative treatment options including dietary modification, physical activity and behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI (> 35 kg/m2). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status.
Roux-en-Y gastric bypass (RYGB) is the most effective surgical procedure [@Berthoud_2011]. Here, a small pouch is formed from the proximal stomach and connected to the jejunum, disconnecting the larger part of the stomach and the proximal part of the small intestine intact from the digestive flow. Other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) reduce stomach volume, but leave the small intestine unchanged [@Mulla_2017].
RYGB-induced weight loss is in part mediated by malabsorption of certain nutrients, especially fat, and reduced digestion efficiency [@Berthoud_2011] while the role of increased energy expenditure after RYGB is still unclear [@Yarmush_2017;@Lamarca_2019]. Most importantly, RYGB alters the signaling cascades between the gut, fat tissue and the brain and thereby induces profound changes in the regulation of food intake and energy homeostasis [@Berthoud_2011]. RYBG-patients frequently report altered smell and taste perception, i.e. increased sensitivity to sweet taste, show less desire to consume high-calorie and high-fat foods and actually reduce intake of such food items [@Mulla_2017, @Berthoud_2011].  These changes in appetitive and reward signaling likely relate to altered gut-brain communication, mediated by afferent vagal nerve signaling and gut hormones like GLP-1, PYY and ghrelin [ @Brutman_2019].

Yet, precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated.
Resting state functional magnetic resonance imaging (rsfMRI) is a powerful technique to investigate the dynamic organization of the brain via the correlation of spontaneous neural activity. Anatomically distinct brain regions show functional connectivity (FC), i.e. correlated neural activity over time, in the absence of a cognitive task. Crucially, these brain networks correspond to networks activated during specific cognitive processes, e.g. visual, auditory and somatosensory processing, or higher-order cognitive functions, i.e. attention and executive control [@Smith_2009], and can thus be seen as default configuration of the brain. In the context of bariatric surgery, the default mode network, a higher-order network, involved in interoception and governing shifts between external-internal processes, and the reward network, processing hedonic value and internal motivation, are promising candidates to mediate altered gut-brain signaling.
The default mode network has been discovered as a “task-negative” network, and is involved in a variety of largely internally-directed cognitive functions including interoception, mind wandering and memory. Its core hubs are the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015].
Higher BMI and obesity have been associated with differences in DMN FC in several resting state and task-based fMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015, @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018].
Despite the large heterogenity and low median power, these findings might be summarized as a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks. This has been interpreted as impaired interoceptive processing failing to adequately balance internal and external behavioral cues.
Longitudinal studies largely support this view. Increased activation of the DMN in a food-viewing paradigm, i.e. insufficient down-regulation of DMN, has been shown to correlate with appetite measures after an overnight fast [@Tregellas_2011, @Wijngaarden_2015], an effect that could be partially reversed by an exercise intervention [@McFadden_2013]. In a bariatric surgery study, [@Li_2018] reported a reduction of local FC in mid-line DMN regions in surgical patients compared to an obese control group. Simultaneously, PCC and medial prefrontal cortex were more strongly connected to dorsolateral prefrontal cortex, a region implicated in cognitive control. Another study in bariatric surgery patients showed that heightened connectivity of the PCC with anterior superior temporal gyrus (STG) before the surgery decreased over the course of one year [@Olivo_2017]. Similarly, the profile of DMN connectivity with frontal regions, including ACC, frontal superior gyrus and the right OFC,  was similar in patients after bariatric surgery and normal weight controls, but dissimilar to that of obese controls [@Frank_2013].

Another important brain network in the context of obesity is the reward network, comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of reward network regions during anticipation of (high-caloric) food cues, and in contrast, reduced reward region activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020;]. Food cue hyperactivation has been identified as a main neural vulnerability factor predicting future weight gain [@Stice_2015]. Resting state fMRI studies have partly mirrored these results, showing increased local FC of reward network regions, i.e. Nacc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. Imaging studies in bariatric surgery patients provided first evidence that hyperconnectivity between reward network regions (e.g. putamen, OFC) might be normalized to some extent by the surgery [@Duan_2020; @Wiemerslage_2016]. Possible, the reconfiguration of the fronto-striatal brain networks might be related to altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013, but @Zoon_2018] .
Thus, while there is some evidence hinting to a role for DMN and reward network connectivity in altered gut-brain communication after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohort of patients, often without adequate control group, and employed exploratory statistics [@George_2016].

In this study, we aimed to extend the current literature and investigate differences of DMN and reward network connectivity before and after bariatric surgery in preregistered analyses and in comparison with a contemporaneously measured waiting-control group. Moreover, we explored whether the changes in functional connectivity were associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential dose-response relationship.
Participants were measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we took into account possible non-linear time courses in increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017].

@@

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité Berlin. For more details see @Prehn_2020. We used all data acquired until April 2019.  
The study protocol was in accordance with the Helsinki Declaration and approved by the local ethics committee. The study was registered at clinicaltrials.gov as NCT01554228. We preregistered the present resting state analyses on the [OSF](https://osf.io/yp42s). We made [additional changes](https://osf.io/59bh7/) to the preregistration after preprocessing as we realized some aspects of the analysis were inadequately described in the first preregistration.   
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->
Participants were recruited from the Center for Bariatric and Metabolic Surgery. Inclusion criteria were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants were aged between $18$ and $70$ and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$).
In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see section \@ref{sec:Anatomical Quality control}).
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availabile. In more detail, `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point. In the intervention group, data is not missing completely at random, as the pre-surgery timepoint MRI could not be obtained in some patients due to the limited scanner bore.
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "../report/dfBL.rds")

@@

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
<!-- BASELINE CHARACTERISTICS -->
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and mean framewise displacement (FD), and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).

@@

```{r figBMIdescr, fig.align='center', out.width="100%", fig.cap="Trajectoriy of BMI separately for the bariatric surgery group (BARS) and the waiting-list control group (NBARS); individual trajectories are plotted in transparent, mean trajectories including standard deviations in opaque colors.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
readRDS(file = "../report/fig/fig_BMIdescr.rds")
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropocentric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

@@

#### Functional Quality control
Resting state fMRI data quality control was performed according to the protocol by @Ciric_2018.
As the expected average head motion was high in this sample, it was difficult to determine an appropriate exclusion threshold which would not lead to the exclusion of a high number of individuals [@Beyer_2020]. Thus, we did not exclude any participants based on mean FD estimates beyond those who failed the anatomical preprocessing. We applied best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. In addition, for models showing a significant time-by-group interaction, we performed a sensitivity analysis excluding the 10% data sets with highest mean FD.
```{r carpetplot, out.width="50%", out.height="30%", fig.cap="Carpet plot for quality control", fig.subcap = c("Example 1 with strong amount of structured noise", "Example 2 with less amount of structured noise"), fig.show='hold',fig.align='center', echo=FALSE}
carpets=c("../report/carpetplotadi006fu2.jpeg","../report/carpetplotadi16fu.jpeg")
knitr::include_graphics(carpets)
```
On the single-subject level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref(fig:carpetplot) shows an exemplary carpet plot with (from top to bottom): percent outliers defined by AFNI, mFD, `OutlierCount`, DVARS (spatial root mean square of the data after temporal differencing),voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data (red: GM voxels, green: WM voxels, blue: ventricle, green: cerebellum). All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.
We evaluated mFD-FC relationships by computing a functional connectome based on [@Power_2012]. In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we correlated mean FD and QC values for each node across participants, and plotted the resulting mFD-FC correlation values in histogram (see Figure \@ref{exampleFC}. We reported the mFD-FC correlation per preprocessing scheme.
Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward network over all participants and timepoints, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with N=999. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average connectivity from these masks for GSR regressed data.
*In the preregistration we wrote: "We will use the average of connectivity values within masks of the respective networks as aggregate network connectivity measure. The mask will include all voxels surviving FWE-correction of p<0.05 in a one-sample t-test of the individual connectivity maps across all participants and time points, calculated with the modified SwE option in the SwE toolbox"*

@@

```{r "qa check DMN model age,sex", eval = FALSE}
diagnostics.plot(R0)
ranef.diagn.plot(R0)
